---
title: "NYC SAT Scores Analysis"
author: "Tony Fischetti"
date: "November 1, 2015"
output: html_document
---

#### Aim:
To use Bayesian analysis to compare NYC's 2010 
combined SAT scores against the average of the
rest of the country, which, according to
FairTest.com, is 1509


```{r, echo=FALSE}
# options
options(echo=TRUE)
options(stringsAsFactors=FALSE)
```

We are going to use the `assertr` and `runjags`
packages for data checking and MCMC, respectively.
```{r}
# libraries
library(assertr)   # for data checking
library(runjags)   # for MCMC
```

Let's make sure everything is all set with JAGS!
```{r}
testjags()
```
Great!

This data was found in the NYC Open Data Portal:
https://nycopendata.socrata.com
```{r}
link.to.data <- "http://data.cityofnewyork.us/api/views/zt9s-n5aj/rows.csv?accessType=DOWNLOAD"
download.file(link.to.data, "./data/SAT_Scores_NYC_2010.csv")

nyc.sats <- read.csv("./data/SAT_Scores_NYC_2010.csv")
```

Let's give the columns easier names
```{r}
better.names <- c("id", "school.name", "n", "read.mean",
                  "math.mean", "write.mean")
names(nyc.sats) <- better.names
```

There are `r nrow(nyc.sats)` rows but almost 700 NYC schools. We will,
therefore, *assume* that this is a random sample of NYC schools.


Let's first check the veracity of this data...
```{r, error=TRUE}
nyc.sats <- assert(nyc.sats, is.numeric,
                   n, read.mean, math.mean, write.mean)
```

It looks like check failed because there are "s"s for some rows. (??)
A look at the data set descriptions indicates that the "s" is for schools
with 5 or fewer students. For our purposes, let's just exclude them.


This is a function that takes a vector, replaces all "s"s
with NAs and make coverts all non-"s"s into numerics
```{r}
remove.s <- function(vec){
  ifelse(vec=="s", NA, vec)
}

nyc.sats$n          <- as.numeric(remove.s(nyc.sats$n))
nyc.sats$read.mean  <- as.numeric(remove.s(nyc.sats$read.mean))
nyc.sats$math.mean  <- as.numeric(remove.s(nyc.sats$math.mean))
nyc.sats$write.mean <- as.numeric(remove.s(nyc.sats$write.mean))
```

Now we are going to remove schools with fewer than 5 test takers
and calculate a combined SAT score
```{r}
nyc.sats <- nyc.sats[complete.cases(nyc.sats), ]

# Calculate a total combined SAT score
nyc.sats$combined.mean <- (nyc.sats$read.mean +
                           nyc.sats$math.mean +
                           nyc.sats$write.mean)
```

Let's now build a posterior distribution of the true mean
of NYC high school's combined SAT scores. We're not going to look
at the summary statistics because we don't want to bias our priors.
We will use a standard gaussian model.

```{r, cache=TRUE, results="hide", warning=FALSE, message=FALSE}
the.model <- "
model {
  # priors
  mu ~ dunif(0, 2400)
  stddev ~ dunif(0, 500)
  tau <- pow(stddev, -2)

  # likelihood
  for(i in 1:theLength){
     samp[i] ~ dnorm(mu, tau)
  }
}"

the.data <- list(
  samp = nyc.sats$combined.mean,
  theLength = length(nyc.sats$combined.mean)
)

results <- autorun.jags(the.model, data=the.data,
                        n.chains = 3,
                        monitor = c('mu'))
```

Let's view the results of the MCMC.
```{r}
print(results)
```

Now let's plot the MCMC diagnostics
```{r, message=FALSE}
plot(results, plot.type=c("histogram", "trace"), layout=c(2,1))
```

Looks good!


Let's extract the MCMC samples of the mean and get the
bounds of the middle 95%
```{r}
results.matrix <- as.matrix(results$mcmc)
mu.samples <- results.matrix[,'mu']
bounds <- quantile(mu.samples, c(.025, .975))
```

We are 95% sure that the true mean is between `r bounds[1]` and
`r bounds[2]`.

Now let's plot the marginal posterior distribution for the mean
of the NYC high schools' combined SAT grades and draw the 95%
percent credible interval.
```{r}
plot(density(mu.samples),
     main=paste("Posterior distribution of mean combined SAT",
                "score in NYC high schools (2010)", sep="\n"))
lines(c(bounds[1], bounds[2]), c(0, 0), lwd=3, col="red")
```

Given the results, the SAT scores for NYC high schools in 2010
are **incontrovertibly** not on par with the average SAT scores of
the nation.

------------------------------------

This is some session information for reproducibility:
```{r}
devtools::session_info()
```